---
title: "Exploring Markus' weevil population data"
author: "David Nguyen"
date: "March 17, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(readxl)
```


```{r, include = FALSE}
# get names of sheets that contain pop data (not including T15-8 for now)
sheet_names <- excel_sheets("weevil_data/pop_dynamics_expt_data.xlsx")
sheet_pops <- sheet_names[1:5]
pct_peanut_shells <- c(0, 0.03, 0.05, 0.10, 0.15)

# create new colnames
npop <- 10
date_cols <- c("days", "date")
demo_cols <- unlist(lapply(1:npop,
       function(i)
         c(paste0("m_alive_",i),
           paste0("f_alive_",i),
           paste0("total_alive_", i),
           paste0("m_dead_",i),
           paste0("f_dead_",i),
           paste0("total_dead_",i))))
treatment_cols <- c("m_added", "f_added", "pellets", "noclue", "treatment")

# read-in and reformat all population data sheets

list_dfs <- vector("list", length = length(sheet_pops)) # init list to store each df

# loop to read and reformat
for (i in seq_along(sheet_pops)) {
  df_now <-
    read_excel("weevil_data/pop_dynamics_expt_data.xlsx",
               skip = 3,
               sheet = sheet_pops[i]) %>%
    mutate(treatment = pct_peanut_shells[i])
  
  # assign new colnames
  names(df_now) <-
    c(date_cols, demo_cols, treatment_cols) # should include check to make sure all data sheets have same format
  
  # T15 has weird ??? in a col. CHange to NA and convert col to numeric
  if (i == 5) {
    df_now$m_alive_8 <-
      df_now %>% 
      mutate_at(vars('m_alive_8'), ~na_if(., '???')) %>%
      pull(m_alive_8) %>%
      as.numeric()
  }
  
  # make data longer for plotting
  df_long <- df_now %>%
    select(-pellets,-noclue) %>% # always 2 pellets a day, and drop noclue as well as the num. added
    filter(!is.na(days)) %>%  # drop empty bottom two rows (included when reading due to sum(pellets) in spread sheet)
    pivot_longer(
      -c(days, date, m_added, f_added,treatment),
      names_sep = "_",
      names_to = c("sex", "status", "replicate"),
      values_to = "count"
    ) %>%
    # original data file did not have totals computed for all days
    # create new df with totals for all days
    filter(sex != "total" , status == "alive") %>%
    group_by(date, replicate) %>%
    mutate(total = sum(count, na.rm = TRUE)) %>% # females were not observed after first few days
    ungroup()
  
  # save to list
  list_dfs[[i]] <- df_long
}

full_df <- bind_rows(list_dfs)

full_df <- full_df %>% mutate(replicate = as.numeric(replicate))
#full_df$treatment <- as_factor(full_df$treatment) # for proper facetting when plotting. Should convert to numeric instead.
```

Markus collected an incredible amount of data (100 population time series, most at least 400 days).

This is obviously great data to practice fitting ecological models (and will provide more practice data munging than I would have liked). Depending on whether we can resolve some of the data quality issues I've found, it may be possible to do some publishable research with this.

There was actually a recent paper in Ecology that did what it sounds like Markus was trying to do. They used early warning signals to try to predict if populations were going to crash. They found that information on traits (here, body size) improved the reliability of EWS for predicting population crashes in simulated data and microcosms. I wonder if we can use these data to build on their approach or do something complimentary?

paper: [Effect of time‐series length and resolution on abundance‐ and trait‐based early warning signals of population declines](https://esajournals.onlinelibrary.wiley.com/doi/abs/10.1002/ecy.3040)

### plot time series from "pop dynamics expt data"

```{r}
full_df %>%
  ggplot(aes(x = days, y = total, group = replicate, col = as.factor(replicate) )) +
  geom_line() + 
  facet_wrap(~treatment) +
  theme_light() +
  theme(legend.position = "none") +
  scale_x_continuous(breaks = seq(0, 650, by = 100)) +
  scale_y_continuous(breaks = seq(0, 50, by = 10)) +
  labs(title = "Bean weevil populations over time",
       subtitle = "For each plot, weevils were fed pellets with a different proportion of peanut shell additive.\nEach color is a different replicate.",
       y = "Bean weevil count",
       x = "Time (days)")

```

Interesting. So, it appears that at 0 - 3 % peanut shells, the weevils can reach a carrying capacity with maybe stochastic or periodic fluctuations. At 5 % peanut shells, they *appears* (at least to my eye) to create more extreme oscillilations than in ithe 0 - 3 % shell treatments which brought a few populations to extinctinction and many dangerously close to extinction at population troughs. For the 10 - 15 % peanut shell treatments the trajectories were clearly declining from the start but had some oscillations before crashing.

Alaina had a good observation that the 5 % population seemed to have bigger transient peaks than the pure bean populations. She thought maybe this is due to compensatory reproduction by the weevils. She reasoned  that since all the weevils were probably raised on pure beans (need to check this) that maybe when they were put on the 5 % beans that the spikes in population were due to overcompensating fecundity as a stress response to poor quality habitat. I wonder if there is a signal of this in the populations where food quality was slowly decreased over time (or maybe the gradual change wouldn't trigger the stress response?). 

The 3 % shell experiment is short because it was started at a later date than all other experiments but was ended at the same date. I'm guessing Markus noticed the huge difference in dynamics between 0 and 5 % and wanted to check the behavior intermediate to 0 and 5 %. It seems that 3 % was more like the 0 % treatment than the 5 % treatment.

```{r eval = FALSE}
full_df %>%
  ggplot(aes(x = date, y = total, group = replicate, col = as.factor(replicate) )) +
  geom_line() + 
  facet_wrap(~treatment) +
  theme_light() +
  theme(legend.position = "none") +
  #scale_x_continuous(breaks = seq(0, 650, by = 100)) +
  scale_y_continuous(breaks = seq(0, 50, by = 10)) +
  labs(title = "Bean weevil populations over time",
       subtitle = "For each plot, weevils were fed pellets with a different proportion of peanut shell additive.\nEach color is a different replicate.",
       y = "Bean weevil count",
       x = "Time (days)")

```

Here's a closer look at the different time series.

```{r}
# increments for x and y axis
inc_total <- 5
inc_days <- 50 

# get max days and total, each rounded to the nearest value of inc_*
spark_lims <- 
  full_df %>% 
  select(days, total) %>% 
  summarise(max_days = ceiling(max(days)/inc_days)*inc_days, 
            max_total = ceiling(max(total)/inc_total) * inc_total )

spark_list <- lapply(seq_along(pct_peanut_shells),
                     function(x)
                       full_df %>%
  filter(treatment == pct_peanut_shells[x]) %>%
  ggplot(aes(x = days, y = total)) +
  geom_line() +
  scale_y_continuous(breaks = seq(0, spark_lims$max_total, by = 2*inc_total),
                     limits = c(0, spark_lims$max_total)) +
  scale_x_continuous(breaks = seq(0, spark_lims$max_days, by = 2*inc_days),
                     limits = c(0, spark_lims$max_days)) +
  labs(title = paste0(pct_peanut_shells[x]*100, " % peanut shell in food pellets")) +
  facet_wrap(~replicate))

# 0 % peanut shells
spark_list[[1]]
```

```{r}
spark_list[[2]]
```

```{r}
spark_list[[3]]
```

```{r}
spark_list[[4]]
```

```{r}
spark_list[[5]]
```

### phase space
Lets take a look at the timeseries in phase space.

```{r phase_space, eval = FALSE}
# might be useful in the future
# max(full_df$total)

phase_list <- lapply(seq_along(pct_peanut_shells),
                     function(x)
                       full_df %>%
  filter(treatment == pct_peanut_shells[x]) %>%
  ggplot(aes(x = lag(total), y = total)) +
  geom_path() +
   scale_y_continuous(#breaks = seq(0, spark_lims$max_total, by = 2*inc_total),
                      limits = c(0, 55)) +
   scale_x_continuous(#breaks = seq(0, spark_lims$max_days, by = 2*inc_days),
                      limits = c(0, 55)) +
  labs(title = paste0(pct_peanut_shells[x]*100, " % peanut shell in food pellets")) +
  facet_wrap(~replicate))

phase_list[[1]]
phase_list[[2]]
phase_list[[3]]
phase_list[[4]]
phase_list[[5]]
```


### plot time series from "ext data""


```{r read_ext_control, include = FALSE}
# get names of relevant sheets
control_sheets <- excel_sheets("weevil_data/ext_data.xlsx")[1:3]

# read in sheet
# sheets each have replicates 1 - 10, 11 - 20, 21 - 30 
lower_rep <- c(1, 11, 21)
upper_rep <- c(10, 20, 30)

list_ctrl <- vector("list", length = length(control_sheets))

for (j in seq_along(control_sheets)) {
  date_cols <- c("days", "date")
demo_cols_ext <- unlist(lapply(lower_rep[j]:upper_rep[j],
       function(i)
         c(paste0("m_alive_",i),
           paste0("f_alive_",i),
           paste0("total_alive_", i),
           paste0("m_dead_",i),
           paste0("f_dead_",i),
           paste0("total_dead_",i))))
treatment_cols_ext <- c("m_added", "f_added", "pellets", "noclue")

raw_ctrl <- read_excel("weevil_data/ext_data.xlsx", skip = 3, sheet = control_sheets[j]) %>%
  filter(!is.na(date) ) # cut out calculated values at bottom of spreadsheet

names(raw_ctrl) <- c(date_cols, demo_cols_ext, treatment_cols_ext)

# create new cols for all treatments
raw_ctrl <- 
  raw_ctrl %>%
  mutate(treatment = "control",
         pct_shell = 0,
         avidin = 0,
         harvest = 0)

# lengthen data for plotting
list_ctrl[[j]] <-
  raw_ctrl %>%
    select(-pellets,-noclue) %>% # always 2 pellets a day, and drop noclue as well as the num. added
    filter(!is.na(days)) %>%  # drop empty bottom two rows (included when reading due to sum(pellets) in spread sheet)
    pivot_longer(
      -c(days, date, m_added, f_added,treatment, pct_shell, avidin, harvest),
      names_sep = "_",
      names_to = c("sex", "status", "replicate"),
      values_to = "count"
    ) %>%
    # original data file did not have totals computed for all days
    # create new df with totals for all days
    filter(sex != "total" , status == "alive") %>%
    group_by(date, replicate) %>%
    mutate(total = sum(count, na.rm = TRUE)) %>% # females were not observed after first few days
    fill(m_added, f_added) %>% # fill in rest of zeros
    ungroup() 

  
}

ctrl_df <- 
  bind_rows(list_ctrl) %>%
  filter(!is.na(count), sex == "m") %>% # cut out rows after experiment ended (e.g., no values recorded for male count)
  mutate(replicate = as.numeric(replicate) )
```



```{r read_harvest_date, include = FALSE}
# get names of relevant sheets
harvest_sheets <- excel_sheets("weevil_data/ext_data.xlsx")[5:6]


list_harvest <- vector("list", length = length(harvest_sheets))

lower_rep <- c(11, 21)
upper_rep <- c(20, 30)

for (j in seq_along(harvest_sheets)) {
  date_cols <- c("days", "date")

  if (harvest_sheets[j] == "XH11-20") {
    demo_cols_harvest <- unlist(lapply(lower_rep[j]:upper_rep[j],
       function(i)
         c(paste0("m_alive_",i),
           paste0("f_alive_",i),
           paste0("total_alive_", i),
           paste0("m_dead_",i),
           paste0("f_dead_",i),
           paste0("total_dead_",i))))
  } else {
    demo_cols_harvest <- unlist(lapply(lower_rep[j]:upper_rep[j],
       function(i)
         c(paste0("m_alive_",i),
           paste0("f_alive_",i),
           paste0("m_dead_",i),
           paste0("f_dead_",i))))
  }
  treatment_cols_ext <- c("m_added", "f_added", "pellets", "harvest", "noclue")

raw_harvest <- read_excel("weevil_data/ext_data.xlsx", skip = 3, sheet = harvest_sheets[j]) %>%
  filter(!is.na(date) ) # cut out calculated values at bottom of spreadsheet
  
  # remove extra col for rep 21 -30 sheet
  if (harvest_sheets[j] == "XH21-30") raw_harvest <- raw_harvest %>% select(1:47)

names(raw_harvest) <- c(date_cols, demo_cols_harvest, treatment_cols_ext)

# create new cols for all treatments
raw_harvest <- 
  raw_harvest %>%
  mutate(treatment = "harvest",
         pct_shell = 0,
         avidin = 0)

# lengthen data for plotting
list_harvest[[j]] <-
  raw_harvest %>%
    select(-pellets,-noclue) %>% # always 2 pellets a day, and drop noclue as well as the num. added
    filter(!is.na(days)) %>%  # drop empty bottom two rows (included when reading due to sum(pellets) in spread sheet)
    pivot_longer(
      -c(days, date, m_added, f_added,treatment, pct_shell, avidin, harvest),
      names_sep = "_",
      names_to = c("sex", "status", "replicate"),
      values_to = "count"
    ) %>%
    # original data file did not have totals computed for all days
    # create new df with totals for all days
    filter(sex != "total" , status == "alive") %>%
    group_by(date, replicate) %>%
    mutate(total = sum(count, na.rm = TRUE)) %>% # females were not observed after first few days
    fill(m_added, f_added) %>% # fill in rest of zeros
    ungroup() 

  
}

harvest_df <- bind_rows(list_harvest) %>% mutate(replicate = as.numeric(replicate))

```


```{r read_food_date, include = FALSE}
# read in sheet
# sheets each have replicates 1 - 10, 11 - 20, 21 - 30 
lower_rep <- c(1, 11, 21)
upper_rep <- c(10, 20, 30)

# get names of relevant sheets
food_sheets <- excel_sheets("weevil_data/ext_data.xlsx")[7:9]

list_food <- vector("list", length = length(food_sheets))

date_cols <- c("days", "date")

for (j in seq_along(food_sheets)) {
  # note, there are no "TOT" cols in the orignal food data set
demo_cols_food <- unlist(lapply(lower_rep[j]:upper_rep[j],
       function(i)
         c(paste0("m_alive_",i),
           paste0("f_alive_",i),
           paste0("m_dead_",i),
           paste0("f_dead_",i))))

# food treatment cols
treatment_cols_food <- c("m_added", "f_added", "pellets", "pct_shell", "noclue")

raw_food <- read_excel("weevil_data/ext_data.xlsx", skip = 3, sheet = food_sheets[j]) %>%
  filter(!is.na(date)) %>% # cut out calculated values at bottom of spreadsheet
  select(1:47)#-`...48`, -`dead FX for pics`, -`...49`) # remove extraneous cols

names(raw_food) <- c(date_cols, demo_cols_food, treatment_cols_food)

# cut out rows that have calculated values
if (food_sheets[j] == "XF1_10") raw_food <- raw_food %>% filter(days < 456)

# create new cols for all treatments
raw_food <- 
  raw_food %>%
  mutate(treatment = "food",
         # pct_shell = 0,
         avidin = 0,
         harvest = 0)

# change pct_shell col to numeric following food quality change schedule
number_pure <- raw_food %>% filter(pct_shell == "pure") %>% nrow() 
pure_beans <- rep(0, number_pure)
shell_beans <- rep(0.01 * seq(1,17,by = 1), each = 5)
pct_shell_treatment <- c(pure_beans, shell_beans)
remaining_rows <- nrow(raw_food) - length(pct_shell_treatment)

raw_food$pct_shell <- append(pct_shell_treatment, rep(NA, remaining_rows) )

  # m_dead_29 hs "?" in a col. CHange to NA and convert col to numeric
  if (j == 3) {
    raw_food$m_dead_29 <-
      raw_food %>% 
      mutate_at(vars('m_dead_29'), ~na_if(., '?')) %>%
      pull(m_dead_29) %>%
      as.numeric()
  }

# lengthen data for plotting
list_food[[j]] <-
  raw_food %>%
    select(-pellets,-noclue) %>% # always 2 pellets a day, and drop noclue as well as the num. added
    filter(!is.na(days)) %>%  # drop empty bottom two rows (included when reading due to sum(pellets) in spread sheet)
    pivot_longer(
      -c(days, date, m_added, f_added,treatment, pct_shell, avidin, harvest),
      names_sep = "_",
      names_to = c("sex", "status", "replicate"),
      values_to = "count"
    ) %>%
    # original data file did not have totals computed for all days
    # create new df with totals for all days
    filter(sex != "total" , status == "alive") %>%
    group_by(date, replicate) %>%
    mutate(total = sum(count, na.rm = TRUE)) %>% # females were not observed after first few days
    fill(m_added, f_added) %>% # fill in rest of zeros
    ungroup()
  
}

food_df <- bind_rows(list_food) %>% mutate(replicate = as.numeric(replicate))

# food_df %>%
#   ggplot(aes( x = days, y = total, col = as.factor(replicate))) + geom_line()
```

The ext data file contains the majority of the time series Markus collected. He had two treatments:

* linearly increasing harvest on pure beans (20 replicates)

* linearly increasing peanut shell percentage without harvesting (30 replicates)

* controls on pure bean (30 replicates)

The treatments populations each had an "acclimation phase" in the first 84 days (22 time points) where no treatment was implemented to allow populations to establish. Then the treatments were ramped up linearly. 

In the acclimation phase, all populations were under the same conditions - I think it is quite remarkable how similar all the populations are. It would be interesting to fit models to different durations of the time series to identify at what point we first notice that something is happening to the population. 

This would be particullarly interesting in the food quality degredation treatments since I think it illustrates a gradual forcing in habitat degredation which is a concern for real populations. These data could allow us to pose and answer the question "how long does it take to detect the impact of habitat degredation on population viability?" 

We could also use the constant food quality experiments to figure out if we can predict the time to extinction or quasi-extinction in these decreasing food quality experiments is predictable based on our knowledge of populations in constant food conditions. This could help us understand the use of short time series of population dynamics across a range of habitat quality for determining population feasibility in locations where habitat continues to degrade or to identify ecological "safe zones" for population persistence. As a concrete example, lets say that wind farms are built at bird nesting sites. Using short term data on bird population dynamics, can we use just time series (without individual level survival and reproduction data) to determine population viability and identify limits on wind farm operations that allow for population persistence?

```{r}
ext_df <- bind_rows(harvest_df, food_df, ctrl_df)
ext_df %>%
  ggplot(aes(x = days, y = total, col = as.factor(replicate) )) +
  geom_line() +
  geom_vline(aes(xintercept = 84)) +
  theme(legend.position = "n") +
  facet_wrap(~treatment, nrow = 3) +
  labs(title = "Extinction experiment time series",
       y = "weevil abundance",
       subtitle = "Before the vertical line conditions were the same across all replicates")
```


As with data we looked at earlier, populations on pure bean reach a stochastic carrying capacity.

```{r}
ctrl_df %>%
  ggplot(aes(x = days, y = total)) +
  geom_line() +
  facet_wrap(~replicate) +
  labs(title = "control expts")

```


Hmm, the harvest lines were never actually pushed to extinction...
From what I can tell, the harvests were done every 4 days. Depending on the generation times of weevils, we maybe wouldn't theoretically expect them to go extinct under proportional quota harvesting, especially considering that harvesting never exceeded 17 %.

There are only 20 replicates in this treatment because the first ten (Sheet labeled "XH1-10") were not actually harvesting treatment. That sheet had a treatment labelled avidin, which, from cursory googling appears I believe is an alternative way that Markus tried to reduce food quality. (Avidin is a type of glycoprotein that binds biotin. So, when added to stored grains, there is less nutrient bioavailibity for pests like these cowpea weevils. This decreases insect growth). I did haven't cleaned up the avidin data yet, since there are a lot of missing data in that sheet (e.g., missing entries for avidin treatment amount, the units of the avidin treatment...)

```{r}
harvest_df %>%
  ggplot(aes(x = days, y = total)) +
  geom_line() +
  geom_line(aes(x = days, y = harvest), col = "red") +
  facet_wrap(~replicate) +
  labs(title = "Harvest expts",
       y = "abundance or percent harvest",
       subtitle = "red lines denote the percent harvest")
```

The percentage of peanut shells increased (i.e., food quality decreased) over time in the food quality experiments. These populations actually crashed unlike the harvested populations. 
<!-- It would be interesting to see if we can use parameter estimates from the constant food quality experiments to predict how weevils responsed to this increasing decline in food quality. Could be relevent for people that are trying to study how populations respond to climate change or habitat degredation. -->

Hmm, visually it doesn't look like these populations have the same "spikes" at the beginning like the weevils on constant 5 % peanut shell did. This could mean that those spikes were nothing, or that they only happen when weevils experience a sudden change in bean quality. Not sure...

```{r}
food_df %>%
  ggplot(aes(x = days, y = total)) +
  geom_line() +
  geom_line(aes(x = days, y = pct_shell * 100), col = "red") +
  facet_wrap(~replicate) +
  labs(title = "Food quality expts",
       y = "abundance or percent peanut shell",
       subtitle = "red lines denote the percent peanut shell in beans")

```

### Data quality issues

### Sheet 3 in "pop dynamics expt data"
Sheet 3 includes a limited duration of count data for each of the replicates. However, the data shown in Sheet 3 doesn't seem to match up with the data in the treatment specific sheets. Markus made two plots in Sheet 3, one that was the average population abundance for the different treatments and the second had each replicate plotted seperately. Here, I reproduce the two plots using either the data shown in Sheet 3 or the data from the treatment specific sheets in the "pop dynamics expt data.xlsx"

```{r, include = FALSE}
sheet3 <- 
  read_excel("weevil_data/pop_dynamics_expt_data.xlsx", sheet = 10, n_max = 21, skip = 2) %>%
  select(1:39) %>% # bunch of stuff in cols past 40, no clue what it is. Is not used in plots on sheet 3
  slice(-1) 

sheet3_names <- names(sheet3)

sheet3_names <- gsub("^\\...", NA, sheet3_names) %>% # if string begins with "..." replace it with NA
  tibble::enframe() %>%
  fill(value) %>%
  group_by(value) %>%
  mutate(index = row_number(),
         name = paste0(value, "_", index)) %>%
  pull(name)

names(sheet3) <- sheet3_names

sheet3 <-
  sheet3 %>%
  mutate(time_step = row_number()) %>%
  pivot_longer( -time_step, # contains("_"),
               names_sep = "_",
               names_to = c("treatment", "replicate"),
               values_to = "total") %>% 
  mutate(treatment = replace(treatment, treatment == "pure", "0.0")) %>%
  mutate(treatment = as.numeric(treatment)) 
    
```


It is obvious that the data are not the same. Sheet 3 shows a transient peak around day ten for the 0.05 and pure treatments which is absent in the treatment specific plot. Also, in sheet 3, there is a bunch of what appears to be abundance data that is not labeled and was not plotted. I have no clue what that data is or where it is from.

```{r}
# plot of average total in sheet 3
sheet3 %>%
  group_by(time_step, treatment) %>%
  summarize(mean_count = mean(total)) %>%
  ungroup() %>%
  ggplot(aes(x = time_step, y = mean_count, group = treatment, col = as.factor(treatment))) +
  geom_line() + theme(legend.position = "n") +
  labs(title = "Sheet 3 - mean population count",
       y = "mean count across replicates",
       x = "time step (4 days?)")

full_df %>%
  group_by(treatment, days) %>%
  summarize(mean_count = mean(total)) %>%
  ungroup() %>%
  filter(days <= 4 * 19, treatment != 0.03) %>%
  ggplot(aes(x = days, y = mean_count, col = as.factor(treatment), group = treatment)) + 
  geom_line() + theme(legend.position = "n") +
  labs(title = "mean population count from treatment sheets")

```

```{r}
sheet3 %>%
  ggplot(aes(x = time_step, y = total, col = as.factor(treatment), group = replicate)) +
  geom_line() +
  facet_wrap(~treatment) +
  labs(title = "Sheet 3 - population count",
       y = "weevil count",
       x = "time step (4 days?)") +
  theme(legend.position = "n")

full_df %>%
  filter(days <= 4 * 19, treatment != 0.03) %>%
  ggplot(aes(x = days, y = total, col = as.factor(treatment), group = replicate)) + 
  geom_line() +
  facet_wrap(~treatment) +
  theme(legend.position = "n") +
  labs(title = "population count from treatment sheets")
  
```


```{r include = FALSE}
# get data from T15-8 and insert into the main data set
t15_8 <- read_excel("weevil_data/pop_dynamics_expt_data.xlsx",
               skip = 3,
               sheet = 6) %>%
  mutate(treatment = 0.15)

demo_names_8 <- c(paste0("m_alive_",8),
                  paste0("f_alive_",8),
                  paste0("total_alive_", 8),
                  paste0("m_dead_",8),
                  paste0("f_dead_",8),
                  paste0("total_dead_",8))
names(t15_8) <- c(date_cols, demo_names_8, treatment_cols)

t15_8 <- 
  t15_8 %>%
  select(-pellets,-noclue) %>% # always 2 pellets a day, and drop noclue as well as the num. added
    filter(!is.na(days)) %>%  # drop empty bottom two rows (included when reading due to sum(pellets) in spread sheet)
    pivot_longer(
      -c(days, date, m_added, f_added,treatment),
      names_sep = "_",
      names_to = c("sex", "status", "replicate"),
      values_to = "count"
    ) %>%
    # original data file did not have totals computed for all days
    # create new df with totals for all days
    filter(sex != "total" , status == "alive") %>%
    group_by(date, replicate) %>%
    mutate(total = sum(count, na.rm = TRUE)) %>% # females were not observed after first few days
    ungroup()

setdiff(names(t15_8), names(full_df))

failed_8_range <-
  full_df %>% 
  filter(treatment == 0.15, replicate == 8, total > 0) %>% 
  pull(date) %>%
  range()

t15_8_range <- t15_8 %>% 
  filter(total > 0) %>% 
  pull(date) %>% 
  range()

subset_8 <- 
  full_df %>% 
  filter(treatment == 0.15,
                   replicate == 8,
                   date <= failed_8_range[2],
                   date >= failed_8_range[1]) %>%
  mutate(which_8 = "1st replicate 8")

t15_8 <- t15_8 %>% mutate(which_8 = "2nd replicate 8")  

```

### what is the difference between replicate 8 in sheet 5 (15%) and sheet 6 ("T15-8") from "pop dynamics expt data"
In sheet 5 replicate 8 was censored (filled in black, but there was still some data in the cols. The last entry for male count was "???"). There is a seperate sheet that is labeled as the 15 % treatment replicate 8. However, the data is clearly different and they were run at different dates.

Presumably something weird happened to the orignal 8th replicate that caused Markus to discard it. However, some thing weird about the "new" 8th replicate is that the weevil additions (that happen at the beginning of all populations) happened 8 times instead of 5 times like all other replicates. Was this just a typo, or was there some reason that this was done?

```{r}
rbind(t15_8, subset_8) %>% 
  ggplot(aes(x = date, y = total, col = which_8)) +
  geom_line() +
  #theme(legend.title = element_blank()) +
  theme(legend.position = "n") +
  labs(title = "comparison of the different 15 % replicate 8 data",
       subtitle = "The first rep. 8 was abandoned (the final data entry is '???')\nThe second rep. 8 was started soon after, but recieved 8 instead of 5 weevil additions")

```



### Rambling questions and thoughts about data

* Are females unobserved or are the counts missing? 
    + for all the replicates, female counts are missing past the first or first few time points
    + The sheet "T 10 sex ratios" includes females counts for the 10 % peanut shell experiment that are not included in the "T10" sheet.
    + Also, the language in the harvest schedule makes it sound like females *are* countable, since it describes harvesting and equal number of males and females. So where is this data?
    + This is problematic, since population growth is likely more limited by females than by males.


<!-- * now that i've pulled all data sets together, how can I automate data quality checks? no clue. need to learn how. -->

* Population T15-8 has 8 weevil population introductions compared to the 5 introduction in other replicates in T15-8. Is this correct?

<!-- In the sheets, some of the cols for "alive" have a highlighted header cell that says "pic". What does this mean? were the counts obtained by looking at a picture? -->

* dropped data columns
    + Dropped the column at the end that has the repeating sequence {a,b,c,d}? I think it has something to do with "sampling day (envelope)" which is the first column in the "schedule" sheet of the "ext data" file.
    + I dropped the pellet count column since, in all expt, weevils were feed 2 pellets per observation day.
    + I've also dropped counts of dead individuals. This may be important information to keep since it would allow us to check if these time series are a census of (at least male) weevils.

**T15-8**
what going on with that expt? in the blacked out cells, there are some data but the last male alive count is "???"
Edit - The data in the T15-8 sheet is different than the blacked-out columns in in the "15 %" sheet. The start date of the T15-8 replicate is different between sheets. I *think* that there was some sort of experimental problem with the original replicate 8 for the 15 % treatment, and that the T15-8 is a new replicate line. This explains why the start date in Dec 1st whereas the start dates for all other replicatees in the 15 % treatment began on Oct 25th

**ext data**
lots of missing data values in treatment cols. Because of this uncertainty, not even going to look at avidin treatment.
Also, there are missing values in harvest rates, not sure if I should impute missing values or leave as NA? Also, harvest rates are not consistent with the harvesting schudule laid out in the "schedule" sheet. In the schedule, harvests should start at 0.5 % and increase by 0.5 % every "subsequent sampling day" (does this mean every four days? this doesn't seem correct). In the data sheet there is a beginning period of no harvest, then it increases by 1 % every 20 days (5 observations).

<!-- **data formatting** -->
<!-- to combine the extinction expt data and the pop dynamics data, I will need to change grouping variables. -->
<!-- There are a couple types of population lines: -->
<!-- + peanut shell (constant or increasing) -->
<!--     ++ there are two sets of 1:10 replicates, one that has constant peanut shell and the other is increasing -->
<!--     ++ how should I account for this when merging data together? Ah, maybe adjust the treatment col to have elements in {food_cst, food_inc, harvest, avidin}. Then I could have seperate numeric cols like: pct_shell, harvest_rate, avidin. -->
<!-- + harvested (increasing) -->
<!-- + avidin treated (increasing) -->

<!-- Note, right now I'm dropping all of the counts of dead individuals. It may be important information to keep though. -->
