---
title: "weevil life history data"
author: "David Nguyen"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(readxl)
```


```{r, include = FALSE}
#  try to do the same for "pure f" sheet, which (i think?) has life history data for all treatments
# it does have > 800rows... so maybe it is all of them?
raw_pure_f <- read_excel("weevil_data/LH data records.xlsx",
           sheet = "pure f") %>%
  # this filter remove a ton of rows (i think because many of the laid eggs never emerge?)
  # need to check this carefully
  filter(!is.na(egg.laid), # drop any rows without date for egg laying, emergence, or adult death
           !is.na(emerged),
           !is.na(adult.dies),
         gender %in% c("f","m")) # sometimes has "2 ind" or NA

# # fix error in data sheet. change 1900-01-25 to 2011-01-25
wrong_date <- raw_pure_f %>% pull(egg.laid) %>% unique() %>% min(na.rm = TRUE)
# correct_date <- wrong_date$year + 111
# temp <- as.POSIXlt(wrong_date)
# temp$year <- as.POSIXlt(wrong_date)$year + 111 # change 1900 into 2011 (technically, we're adding 0 + 111 here, since 1900 is indexed as 0)
raw_pure_f <-
  raw_pure_f %>% mutate(egg.laid = case_when(egg.laid == wrong_date ~ as.POSIXct("2011-01-25"), # wrong timezone (CST instead of UTC)
                                           egg.laid != wrong_date ~ egg.laid),
                      egg.laid = as.Date(egg.laid))
                         

# emerged and adult.dies were read in as characters, need to fix to remove remaining NAs
lh_data <- 
  raw_pure_f %>%
  # correct format of dates
  mutate(egg_date = egg.laid,#already in correct format
         emerge_date = janitor::excel_numeric_to_date(as.numeric(emerged) ),
         death_date = janitor::excel_numeric_to_date(as.numeric(adult.dies) )) %>%
  # standardize bean types
  mutate(Beantype = case_when(Beantype == "pure" ~ "0",
                              Beantype != "pure" ~ Beantype),
         pct_shell = 100 * as.numeric(Beantype)) %>%
  select(ID, pct_shell, gender, egg_date, emerge_date, death_date)

lh_data %>%
  filter(!is.na(death_date),
         death_date != as.Date("1927-05-17") ) %>%
  mutate(lifetime = as.numeric(death_date - emerge_date),
         max_lifetime = max(lifetime, na.rm = TRUE),
         dead_time = (max_lifetime - lifetime)) %>%
  rowwise() %>% 
  # Make LDLDLD format date (http://www.phidot.org/software/mark/docs/book/pdf/chap16.pdf)
  # c(1,0)s for all days alive, 
  # c(1,1) for death, and 
  # c(0,0)s to censored days post-death (up to max_lifetime)
  # then paste together and collapse into a string
  # https://stackoverflow.com/questions/2098368/concatenate-a-vector-of-strings-character
  mutate(encounter_hist = paste(c(rep(c(1,0), times = lifetime), # days alive
                                c(1,1), # death day
                                rep(c(0,0), times = (max_lifetime - lifetime))), # days censored
                                collapse = ",")) %>%
  select(ID, 
         encounter_hist, 
         gender, pct_shell) %>%
  # make into inp format
  mutate(ch = str_replace_all(encounter_hist, pattern = ",", replace = ""),
         comment_id = paste("/*",ID,"*/", sep = ""),
         freq = 1,
         sex = case_when(gender == "f" ~ 0,
                         gender == "m" ~ 1),
         end_line = ";") %>%
  # inp format
  # /* ID */ capture history, freq of observed history, covariates;
  select(comment_id, ch, freq, sex, pct_shell, end_line) %>%
  filter(pct_shell != 20) %>% # drop data since on 2 observations at 20% shell
    write_delim(path = "weevil_data/kf_pure_f.inp",
            col_names = FALSE) # by default, write_delim makes a " " delimited txt file

# turns out that the "RMark" style is exactly what I already had! 
# make an INP anyway so that I could run analysis again in MARK to double check
RMark::convert.inp("weevil_data/kf_pure_f.inp",
                   covariates = c("sex", "pct_shell"),
                   use.comments = TRUE) # uses comments as row names
```

### Weevil longevity data

```{r longevity_plot, out.width="50%", fig.show='hold'}
surv_data <-
  lh_data %>%
  filter(!is.na(death_date),
         death_date != as.Date("1927-05-17") , # weird stuff in this row of data besides date typo
         pct_shell != 20) %>%
  mutate(lifetime = as.numeric(death_date - emerge_date), 
         daily_death = 1/lifetime,
         daily_surv = 1 - daily_death) %>%
  group_by(pct_shell, gender) %>%
  mutate(mean_lifetime = mean(lifetime),
         mean_surv = mean(daily_surv))

# life time plot
surv_data %>%
  ggplot() +
  geom_jitter(aes(x = pct_shell, y = lifetime, col = gender, shape = gender), size = 2, alpha = 0.5) +
  #geom_point(aes(x = pct_shell, y = mean_lifetime, col = gender), shape = 4, size = 2.5, stroke = 3) +
  geom_boxplot(data = surv_data,
              mapping = aes(x = pct_shell, y = mean_lifetime, 
                            group = interaction(pct_shell, gender), fill = gender ), # use interaction to group by multiple vars
              width = 0.75, position = position_dodge(0.25), outlier.shape = NA) +
  ylim(0, 20) +
  labs(title = "Weevil lifetime",
       #subtitle = "X's are means",
       x = "Percent peanut shell in bean",
       y = "lifetime (days)") +
  theme(legend.position = "n")

# daily survival plot
surv_data %>%  
  ggplot() +
  geom_jitter(aes(x = pct_shell, y = daily_surv, col = gender, shape = gender), size = 2, alpha = 0.5) +
  # geom_point(aes(x = pct_shell, y = mean_surv, col = gender), shape = 4, size = 2.5, stroke = 3) +
  #   ylim(0, 1) +
  labs(title = "Weevil daily survival probability",
       subtitle = "Survival probabilities calculated assuming exponentially distributed lifetimes.",
       x = "Percent peanut shell in bean",
       y = "Daily survival probability") +
  geom_boxplot(data = surv_data,
              mapping = aes(x = pct_shell, y = daily_surv, 
                            group = interaction(pct_shell, gender), fill = gender ), # use interaction to group by multiple vars
              width = 0.75, position = position_dodge(0.25), outlier.shape = NA)
  #geom_point(aes(x = pct_shell, y = mean_lifetime, col = gender), shape = 3, size = 2.5, stroke = 2)
```

### Kaplan-Meier survival analysis

All models assume constant survival probability. When I tried to estimate time specific survival probabilities but I got S(t) = 1 for all time points. Not sure what the issue is there.

```{r run_known_fate, include = FALSE}
library(RMark)
weevil_kf <- convert.inp("weevil_data/kf_pure_f.inp",
                   covariates = c("sex", "pct_shell"),
                   use.comments = TRUE) # uses comments as row names

weevil.Mark <- function () {
  
  S.dot <- list( formula = ~ 1, link = "logit" )                  # S(.) constant survival
  S.shell <- list( formula = ~ pct_shell, link = "logit" )        # S(percent peanut shell)
  S.sex <- list(formula = ~ sex, link = "logit")                  # S(sex of weevil)
  S.sexshell <- list(formula = ~ sex + pct_shell, link = "logit") # sex-specific intercepts and same slope
  S.sexshellfull <- list(formula = ~ sex * pct_shell, link = "logit") # sex specific intercepts and slopes
  S.time <- list(formula = ~ time, link = "logit")
  # S.timeI <- list(formula = ~ I(time), link = "logit")            # try to make time
  
  weevil.models <- create.model.list( "Known" )
  
  weevil.fits <- mark.wrapper( model.list = weevil.models, 
                                 data = weevil_kf, model="Known" )
  return( weevil.fits )

}

weevil.results <- weevil.Mark()

```

<!-- should add plot of regression coefficients for each model -->

```{r surv_model_comparison}
weevil.results$model.table %>%
  select(model, AICc, DeltaAICc, weight, Deviance, npar) %>%
  rename("Model" = model, "Weight" = weight, "Parameters" = npar) %>%
  remove_rownames() %>%
  #tibble()
  knitr::kable(digits = 2)
```


```{r get_betas, include = FALSE}
# lapply( weevil.results[ 1:4 ],
#         function( .mark )                        # I think .mark grabs weevil.results[i]
#           .mark$results$beta %>%                 # Parameter estimates
#           rownames_to_column("coefficient") %>%  # make col with name of coefficient
#           as_tibble(rownames = NA) %>%           
#           mutate(model = .mark[[4]])) %>%        # weevil.results[[i]][4] gets name of model
#   bind_rows() 

# second try using a for loop
# beta_list <- vector("list", length = length(model_names))
# 
# for (i in seq_along(model_names)) {
#   model <- model_names[i] # this bit is not working how i want it to
#   beta_list[[i]] <- weevil.results$model$results$beta %>%
#   rownames_to_column("coefficient") %>%
#   as_tibble(rownames = NA) %>%
#   mutate(model = model_names[i])
# }

# get names of models used by rMark
rmark_names <- names(weevil.results) %>% head(5) 
# create named vector of model names I want to use
model_names <- c("S.dot" = "survival ~ intercept"
                 ,"S.sex" = "survival ~ sex"
                 ,"S.sexshell" = "survival ~ sex + shell"
                 ,"S.sexshellfull" = "survival ~ sex * shell"
                 ,"S.shell" =  "survival ~ shell")

# init a list to store each df of beta
beta_list <- vector("list", length = length(model_names))
# extract df of betas for each model and name it
for (i in seq_along(rmark_names)){
  beta_list[[i]] <- 
    weevil.results[[rmark_names[i]]]$results$beta %>% 
    rownames_to_column("coefficient") %>%  # make col with name of coefficient
          as_tibble(rownames = NA) %>%           
    mutate(model = model_names[rmark_names[i]])
}

# reate df of betas
beta_df <-
  beta_list %>% bind_rows() %>%
  mutate(coefficient = case_when(coefficient == "S:(Intercept)" ~ "int.",
                                 coefficient != "S:(Intercept)" ~ str_remove(coefficient, "S:")))

```


```{r param_est_plots}
# beta_sexshell <- 
#   weevil.results$S.sexshell$results$beta %>%
#   rownames_to_column("coefficient") %>%
#   as_tibble(rownames = NA) %>%
#   mutate(model = "survival ~ sex + pct_shell")
# 
# beta_sex <-
#   weevil.results$S.sex$results$beta %>%
#   rownames_to_column("coefficient") %>%
#   as_tibble(rownames = NA) %>%
#   mutate(model = "survival ~ sex")
# 
# rbind(beta_sexshell, beta_sex) %>%
beta_df %>%
  ggplot() +
  geom_point(aes(x = coefficient, y = estimate, group = interaction(coefficient, model)),
             size = 2) +
  geom_errorbar(aes(x = coefficient, ymin = lcl, ymax = ucl, group = interaction(coefficient, model)),
                width = 0, size = 1) +
  geom_hline(aes(yintercept = 0), linetype = 2) +
  facet_wrap(~model, scales = "free_y", nrow = length(model_names) ) +
  coord_flip() +
  labs(x = element_blank(),
       y = "logit scale parameter estimates (95 % CI)",
       title = "model parameter estimates") +
  theme(legend.position = "n")

```

The top two models are Survival ~ Sex and Survival ~ Sex + food quality. But, I don't think there is a real effect of food quality on adult survival. After accounting for adult sex, food quality has almost zero influence on adult survival. Probably driven solely by a few influential points that popped up for the lower food quality treatments (see plot of observed daily survival).

```{r surv_estimates, warning = FALSE}
# make data for predictions
pred_data <- expand_grid(sex = c(0, 1),
                    pct_shell = seq(0, 15, by = 5))

# Generate model predictions
# use covariate.predictions (indices = 1) to predict survival
intercept.predict <- covariate.predictions( weevil.results$S.dot, # choose model
                                           indices = 1, data = pred_data )

shell.predict <- covariate.predictions( weevil.results$S.shell, 
                                       indices = 1, data = pred_data )

sex.predict <- covariate.predictions( weevil.results$S.sex, 
                                       indices = 1, data = pred_data )

sexshell.predict <- covariate.predictions( weevil.results$S.sexshell, 
                                       indices = 1, data = pred_data )

sexshellfull.predict <- covariate.predictions( weevil.results$S.sexshellfull, 
                                       indices = 1, data = pred_data )

averaged.predict <- covariate.predictions(weevil.results, # averaged model estimates by weight
                                        indices = 1, data = pred_data)
# combine predictions into a single df
pred_list <- 
  list(intercept.predict$estimates
,shell.predict$estimates
,sex.predict$estimates
,sexshell.predict$estimates 
,sexshellfull.predict$estimates
#,averaged.predict$estimates)
)

names(pred_list) <- c("survival ~ intercept"
                      , "survival ~ shell"
                      , "survival ~ sex"
                      , "survival ~ sex + shell"
                      , "survival ~ sex * shell"
                      #, "averaged"
                      )

pred_plot <- do.call(rbind, lapply (names(pred_list), function(x) data.frame(pred_list[[x]], model = x)) ) %>%
  select(model, sex, pct_shell, estimate, lcl, ucl) %>%
  as_tibble()

# Create plot of model predictions, with 95% CI
ggplot(pred_plot, aes(x = pct_shell, y = estimate, group = sex, linetype = factor(sex))) +#, col = sex, linetype = as.character(sex) )) +
  geom_line(size = 1.5) +
  ylim(0.75,1) +
  geom_ribbon(aes(x = pct_shell, ymin = lcl, ymax = ucl, group = sex), alpha = 0.5) +
  facet_wrap(~model) +
  labs(title = "Known-fate survival estimates",
       subtitle = "Females are solid lines, males are dashed",
       x = "Percent peanut shell in bean",
       y = "Estimated daily survival (95 % CI)") +
  theme(legend.position = "n")
```

<!-- Should make two plots: 1. predicted lifespan + obs. data; 2. survival prob + obss .data. much more concise. -->

Same plot but with the observed data overlaid onto estimates (outliers removed). Useful or too busy?

```{r warning = FALSE}
# per above comment, trying to add observed data to predicted survival
ggplot() +
  geom_line(data    = pred_plot, 
            mapping = aes(x = pct_shell, y = estimate, group = sex, linetype = factor(sex) ),
            size = 2) +
  geom_ribbon(data    = pred_plot,
              mapping = aes(x = pct_shell, ymin = lcl, ymax = ucl, group = sex), alpha = 0.5) +
  # geom_jitter(data = surv_data,
  #             mapping = aes(x = pct_shell, y = daily_surv, shape = gender), #, col = gender, shape = gender),
  #             size = 2, alpha = 0.5) +
  geom_boxplot(data = surv_data,
              mapping = aes(x = pct_shell, y = daily_surv, 
                            group = interaction(pct_shell, gender), fill = gender ), # use interaction to group by multiple vars
              width = 0.75, position = position_dodge(0.25), outlier.shape = NA) +
  ylim(0.75,1) +
  # geom_point(data = surv_data,
  #            mapping = aes(x = pct_shell, y = mean_surv, col = gender),
  #            shape = 4, size = 2.5, stroke = 3) +
  #facet_grid(gender~model) +
  facet_wrap(~model) +
  labs(title = "Known-fate survival estimates",
       subtitle = "Females are solid lines or red, males are dashed or blue",
       x = "Percent peanut shell in bean",
       y = "Estimated daily survival (95 % CI)") +
  theme(legend.position = "n")
```

```{r plot_stime, eval = FALSE}
weevil.results$S.time$results$real %>% rowid_to_column("day") %>%
  ggplot() +
  geom_line(aes(x = day, y = estimate)) +
  geom_ribbon(aes(x = day, ymin = lcl, ymax = ucl), alpha = 0.5)
```

The estimates seem reasonable. The estimated survival probabilities are in line with the mean lifetimes shown in the first figure. 

### Thoughts

It is clear that the sex of adult weevils has an effect on survival and that resource quality probably doesn't. 
We should expect that females will have lower survival than males since they allocate more resources to reproduction. 
However, I think it is interesting that resource quality has such a weak effect on survival considering their biology; weevils don't require food after emerging as adults so adult weevils rely almost entirely on what they ingested as larvae. 

Hmm, Markus also recorded the start and end weight of each bean pellet, so we know how much each larvae consumed. 
Do larvae in poor quality beans eat more to compensate for the low nutrition? Not super important, but would be interesting to check out.

The next step is to investigate:

* development time using Kaplan-Meier analysis;
* eclosion success using logistic regression;
* egg laying using poisson regression;

as functions of food quality. 

Once I've done that I'll make population projection matrices with these estimates and compare them with Markus'.

# EDIT: compare constant and age-varying survival models
Note that "time" in these models is synonymous with age (in days), since the times are days since emergence.

It is obvious why the S(t) ~ model fits better than S ~ models. Weevil lifetimes is not well-described by the exponential distribution. This implies that population structure (adult age) may be important to explicitly model.

How do I do time to survival analysis for non-exponentially distributed lifetimes? The KM curve looks concave-down, there is probably a less parameter-rich model for survival times.

```{r calc_St}
# get S(t) for all models.
# need to figure out how to propagate CI forward
pred_plot_km <-
  pred_plot %>%
  group_by(model, sex, pct_shell) %>%
  mutate(day = map(estimate, function(x) 1:20),
         st = map(estimate, function(x) x^(1:20))) %>%
  unnest(cols = c(day, st)) %>%
  select(model, sex, pct_shell, day, st)# %>%
  # mutate(check_st = estimate^day, # double chekc that results are correct
  #        is_wrong = ifelse(st == check_st, FALSE, TRUE)) %>%
  # summarise(not_same = sum(is_wrong)) # everything is fine if not_same == 0

st_pred <-
  weevil.results$S.time$results$real %>% 
  rowid_to_column("day") %>%
  mutate(model = as.factor("S ~ time"),
         sex = as.numeric(NA),
         pct_shell = as.numeric(NA) ) %>%
  select(model, sex, pct_shell, day, st = estimate)


#bind_rows(list(pred_plot_km, st_pred)) %>%
  #pred_plot_km %>%
  ggplot() +
  geom_line(data = pred_plot_km, 
            aes(x = day, y = st, linetype = as.factor(sex), col = as.factor(pct_shell) ),
            size = 1.5) +
    geom_line(data = st_pred,
              aes(x = day, y = st),
              size = 1.5) +
  ylim(0, 1) +
  facet_wrap(~model)

# st_pred %>%
#   ggplot(aes(x = day, y = st)) +
#   geom_line()
```


```{r, eval = FALSE}
# time model
time.predict <- covariate.predictions( weevil.results$S.time, # choose model
                                           indices = 1, data = expand_grid(time = seq(0, 15, by = 1),
                                                                           sex = c(0,1),
                                                                           pct_shell = seq(0, 20, by = 5)) )
time.predict$estimates
```

```{r}
# clean up temp files created by MARK
rm(list = ls())
RMark::cleanup( ask = FALSE )
```


```{r pure_known_fate, include= FALSE, eval = FALSE}
excel_sheets("weevil_data/LH data records.xlsx")
pure_lh_raw <- read_excel("weevil_data/LH data records.xlsx",
           sheet = "all pure data")# %>% 
  # all excel data converted to excel numeric dates
  # mutate to change all back to real dates
pure_lh <- 
  pure_lh_raw %>%
  mutate(egg_date = janitor::excel_numeric_to_date(egg.laid),
         emerge_date = janitor::excel_numeric_to_date(as.numeric(emerged) ),
         death_date = janitor::excel_numeric_to_date(as.numeric(adult.dies) )) %>% 
  select(ID,gender, egg_date, emerge_date, death_date) %>%
  mutate(lifetime = as.numeric(death_date - emerge_date)) 

# create encounter history col for known-fate analysis
kf_pure <- pure_lh %>%
  mutate(max_lifetime = max(lifetime, na.rm = TRUE)) %>%
  rowwise() %>% # rowwise makes var called by mutate row-specific (much handier  than group_by on the row number)
  # in this mutate, i create a vector of 1s for all days alive and 0s to pad days dead (up to max_lifetime)
  # then paste together and collapse into a string
  mutate(encounter_hist = paste(c(rep(1, length.out = lifetime), rep(0, length.out = (max_lifetime - lifetime) )), # https://stackoverflow.com/questions/2098368/concatenate-a-vector-of-strings-character
                                collapse = ","),
         pct_shell = 0) %>%
  filter(!is.na(death_date)) %>% # remove rows where death date is missing (these weevils were sacrificed)
  select(ID, 
         encounter_hist, 
         gender, pct_shell) # include 2 covariates

# what's up with NAs
# i think they are either: never emerged or were sacrificed
pure_lh_raw %>%
  filter(is.na(emerged) | sacrifice == 1) %>%
  select(ID, egg.laid, emerged, adult.dies, sacrifice) %>%
  nrow()
  
```

```{r all_pure_to_INP, eval = FALSE}
# make something to export as inp
kf_pure %>%
  mutate(ch = str_replace_all(encounter_hist, pattern = ",", replace = ""),
         comment_id = paste("/*",ID,"*/", sep = ""),
         freq = 1,
         sex = case_when(gender == "f" ~ 0,
                         gender == "m" ~ 1),
         end_line = ";") %>%
  # inp format
  # /* ID */ capture history, freq of observed history, covariates;
  select(comment_id, ch, freq, sex, pct_shell, end_line) %>%
  #write.table(file= "weevil_data/kf_pure.csv")
  write_delim(path = "weevil_data/kf_pure.inp",
            col_names = FALSE) # by default, write_delim makes a " " delimited txt file

# ughh, turns out that the "RMark" style is exactly what I already had! no point in doing all the work to make an INP format file!
RMark::convert.inp("weevil_data/kf_pure.inp",
                   covariates = c("sex", "pct_shell"),
                   use.comments = TRUE) # uses comments as row names
```

